---
title: "Using because-py via R (reticulate)"
description: |
  How to use the NumPyro engine from R through the because package and reticulate.
  All examples use the bird life-history dataset included in because.phybase.
execute:
  eval: false   # set to true when running in an environment with because + NumPyro installed
---

```{r setup}
#| include: false
library(because)
library(because.phybase)
```

## Overview

The `because` R package calls because-py transparently whenever you specify `engine = "numpyro"`.
You do not need to write any Python code — reticulate handles the bridge.
The workflow is identical to a JAGS run: same formula syntax, same `summary()` and `plot_dag()` methods.

## A simple causal model

```{r simple-model}
library(because)
library(because.phybase)

# Bird life-history data (678 species)
data(birds)

fit <- because(
  equations = list(
    Brain    ~ Mass + Migration,
    Lifespan ~ Brain,
    Clutch   ~ Brain + Mass + Migration
  ),
  data   = birds,
  engine = "numpyro",
  n.iter = 2000,
  n.burnin = 500
)

summary(fit)
```

## D-separation testing first

Following best practice, always test d-separation before comparing models by WAIC.
Only models that pass all d-sep tests are structurally valid.

```{r dsep-first}
# Run d-sep tests without fitting the full model (fast)
fit_dsep <- because(
  equations = list(
    Brain    ~ Mass + Migration,
    Lifespan ~ Brain,
    Clutch   ~ Brain + Mass + Migration
  ),
  data   = birds,
  engine = "numpyro",
  dsep   = TRUE,
  dsep_only = TRUE
)

fit_dsep$dsep
```

## Inspecting the generated NumPyro model code

A unique feature of because-py is that the generated NumPyro model is stored as
an executable Python string, which you can inspect and modify:

```{r inspect-code}
cat(fit$numpyro_code)
```

This is useful for:

- **Verification** — confirming the model matches your theoretical expectations
- **Customisation** — copying the code to a `.py` script and adding custom priors or likelihoods
- **Teaching** — showing students the connection between DAG specification and probabilistic code

## Comparing models

Once d-sep is confirmed, use WAIC to compare models with the same response variable:

```{r compare}
fit_null <- because(
  equations = list(Brain ~ Mass),
  data      = birds,
  engine    = "numpyro",
  n.iter    = 2000
)

because_compare(fit, fit_null)
```

::: {.callout-important}
WAIC comparisons are only valid between models that share the same response variable(s)
and the same observations. Cross-model comparisons with different endogenous variables
will produce inflated WAIC values.
:::

## Extending chains with because_continue()

NumPyro's HMC/NUTS converges so efficiently that `because_continue()` is rarely needed.
If you do need more samples, simply re-run with a larger `n.iter`:

```{r more-samples}
fit2 <- because(
  equations = list(
    Brain    ~ Mass + Migration,
    Lifespan ~ Brain,
    Clutch   ~ Brain + Mass + Migration
  ),
  data     = birds,
  engine   = "numpyro",
  n.iter   = 5000,   # increase from 2000
  n.burnin = 500
)
```

## Visualising the DAG

All plot methods work identically regardless of engine:

```{r plot}
plot_dag(fit)
```

## Traceplots and diagnostics

```{r traceplots}
# Using coda
plot(fit$samples)

# Or with bayesplot
library(bayesplot)
mcmc_trace(fit$samples, pars = c("beta_Brain_Mass", "beta_Brain_Migration"))
mcmc_areas(fit$samples, pars = c("beta_Brain_Mass", "beta_Brain_Migration"))
```
