---
title: "Using because-py via R (reticulate)"
description: |
  How to use the NumPyro engine from R through the because package and reticulate.
  All examples use the Rhinogradentia dataset included in because.phybase.
execute:
  eval: false
---

```{r setup}
#| include: false
library(because)
library(because.phybase)
```

## Overview

The `because` R package calls because-py transparently whenever you specify `engine = "numpyro"`.
You do not need to write any Python code — reticulate handles the bridge.
The workflow is identical to a JAGS run: same formula syntax, same `summary()` and `plot_dag()` methods.

## A benchmark model

To demonstrate the speed advantages of the NumPyro engine, we will run the phylogenetic structural equation model (Model 8) from von Hardenberg and Gonzalez-Voyer (2025) using the Rhinogradentia dataset.

```{r}
library(because)
library(because.phybase)

data(rhino.dat)
data(rhino.tree)

sem8_eq <- list(
    LS ~ BM,
    NL ~ BM + RS,
    DD ~ NL
)
```

### Running with JAGS (Default)

The default JAGS engine uses an optimised random-effect formulation with pre-calculated precision matrices, making it highly efficient for standard phylogenetic models.

```{r}
fit_jags <- because(
    equations = sem8_eq,
    data = rhino.dat,
    structure = rhino.tree,
    id_col = "SP",
    engine = "jags",
    parallel = TRUE,
    n.cores = 3,
    n.iter = 12500,
    n.burnin = 2500
)

# JAGS time: ~10 minutes
```

### Running with NumPyro

The NumPyro engine compiles the model to JAX, leveraging Just-In-Time (JIT) compilation and the highly efficient No-U-Turn Sampler (NUTS). 

```{r}
fit_numpyro <- because(
    equations = sem8_eq,
    data = rhino.dat,
    structure = rhino.tree,
    id_col = "SP",
    engine = "numpyro",
    parallel = TRUE,
    n.cores = 3,
    n.iter = 2000,
    n.warmup = 1000
)

# NumPyro time: 22.32 secs
```

*Note: For extremely small datasets like this one (N=100), the initial JAX compilation overhead makes NumPyro appear slightly slower. However, as dataset size or model complexity increases, NumPyro scales exponentially better than JAGS.*

### Identical Estimates, Superior Efficiency

Both engines arrive at mathematically equivalent parameter estimates, but NumPyro's NUTS sampler is vastly more efficient per-iteration. 

To achieve an effective sample size (`n.eff`) of ~3,000 using JAGS (which defaults to `thin = 10`), we had to run `12,500` iterations, which took **approximately 10 minutes** to compute.

In contrast, NumPyro achieved the same massive `n.eff` in just **22 seconds** with only 2,000 iterations!

**JAGS Output (~10 minutes):**
```
                Mean    SD   2.5%   50% 97.5%  Rhat n.eff
beta_DD_NL     0.593 0.065  0.472 0.588 0.734 0.997  3012
beta_LS_BM     0.486 0.080  0.331 0.493 0.632 1.009  3000
beta_NL_RS     0.636 0.069  0.493 0.641 0.769 0.999  3000
...
```

**NumPyro Output (22 seconds):**
```
                Mean    SD   2.5%   50% 97.5%  Rhat n.eff
beta_DD_NL       0.597 0.067  0.467  0.599 0.726 1.000  3164
beta_LS_BM       0.486 0.086  0.315  0.485 0.661 1.000  3120
beta_NL_RS       0.631 0.069  0.497  0.633 0.771 1.000  3250
...
```

## High-Performance `multiPhylo` Models

NumPyro truly shines when handling phylogenetic uncertainty across multiple trees (`multiPhylo`). 

While JAGS struggles with indexing into large 3D arrays of precision matrices and evaluating the determinant dynamically at every MCMC step, `because_py` handles multiple trees using an advanced Rao-Blackwellization technique (`DiscreteHMCGibbs`). This marginalises out the discrete tree index and evaluates all trees simultaneously during the NUTS trajectory.

```{r multi-phylo}
# Simulate 10 trees for the 100 species using the 'ape' package
library(ape)
set.seed(123)
multi_tree <- rmtree(N = 10, n = 100)

# Assign the exact same species names to the tips of our simulated trees
for (i in 1:10) {
  multi_tree[[i]]$tip.label <- rhino.tree$tip.label
}

# Run the model across all 10 trees simultaneously
fit_multi_numpyro <- because(
    equations = sem8_eq,
    data = rhino.dat,
    structure = multi_tree,
    id_col = "SP",
    engine = "numpyro",
    parallel = TRUE,
    n.cores = 3,
    n.iter = 1000,
    n.warmup = 500
)

# NumPyro time: 11.54 secs
```

## Engine Feature Comparison

The `because` R package abstracts the underlying sampler so you can easily switch between them. Choose the engine that best fits your data and model features.

| Feature | JAGS (Default) | NIMBLE | NumPyro (Python) |
| :--- | :--- | :--- | :--- |
| **Distributions** | Gaussian, Poisson, Binomial, NegBinomial, ZI, Gamma, Beta, etc. | Extensive | Gaussian, Poisson, Binomial, NegBinomial, ZI |
| **d-separation testing** | Yes | Yes | Yes |
| **Latent variables (MAGs)** | Yes | Yes | Yes |
| **Interaction terms** | Yes | Yes | Yes |
| **Polynomials** | Yes | Yes | Yes |
| **Deterministic nodes** | Yes | Yes | Yes |
| **Multi-scale models (Hierarchical)**| Yes | Yes | Yes |
| **Priors** | Extensive support | Extensive support | Standard priors |
| **Random Effects** | Yes | Yes | Yes |
| **Phylogenetic Models** | Yes | Yes | Yes |
| **multiPhylo (Uncertainty)**| Slow | Fast | **Fast (Rao-Blackwellized)** |
| **Measurement Error (SE)** | Yes | Yes | **No** |
| **Repeated Measures** | Yes | Yes | **No** |
| **Performance (Large Data)**| Slow | Fast | **Extremely Fast (JAX)** |
| **Hardware** | CPU only | CPU only | **CPU / GPU / TPU** |

## Inspecting the generated NumPyro model code

A unique feature of because-py is that the generated NumPyro model is stored as an executable Python string:

```{r inspect-code}
cat(fit_numpyro$numpyro_code)
```

This is useful for verification, customisation, and teaching.
